---
title: "Exploring the real-world data in WWHYPDA"
author: "Falk Heße, Nura Kawa"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore WWHYPDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The \pkg{exPrior} packages provides real-world geostatistical data from the World-Wide HYdrological Parameters DAtabank wwwhypda as well as methods for accessing these data using R with the functions \texttt{viewInfo()} and \texttt{getData()}. In this vignette, we are going to investigate these data.

First, we should load the libraries to get access to this function:

```{r}
library(exPrior)
library(ggplot2)
library(dplyr)
library(cowplot)
```

Now we can import the data from WWHYPDA using the internal `getData` function

```{r}
df <- exPrior::getData()
```

After that, we check whether the import was successful

```{r}
dim(df)
```

The dimensions should be 20523 and 39, since this is the full WWHYPDA database.

Now let us create a data frame for eventually plotting some of the data

```{r}
df_porosity <- df %>%
  filter(param_name == 'porosity') %>% # select only measurements of porosity
  filter(site_id %in% c('7', '8', '98', '100', '101', '102')) %>% # select these site IDs
  select(site_id, val) %>% # select the columns site_id and val
  mutate(site_id = factor(site_id,
                          levels = c('7', '8', '98', '100', '101', '102'))) # make site_id a factor
```

To that end, we select only data on porosity from a number of selected sites and make a kernel density plot

```{r}
kde_plot <- ggplot(df_porosity,
                   aes(x = val,
                       fill = site_id)) +
            geom_density() +
            facet_grid(site_id ~ .) + 
            ggtitle('Kernel Density Estimates of Porosity in Six Sites')
kde_plot
```

Now, let us select the hydraulic condictivity 

```{r}
df_hc <- df %>%
  filter(param_name == 'hydraulic conductivity') %>% 
  filter(rt_name %in% c('Basalt',
                        'Chalk',
                        'Coal',
                        'Greensand',
                        'Limestone')) %>%
  select(rt_name, val) %>%
  mutate(rt_name = factor(rt_name))
```

and make a histogram 

```{r}
histogram_plot <- ggplot(df_hc,
                         aes(x = log10(val),
                             fill = rt_name)) +
  geom_histogram(alpha=0.7) +
  xlab('log(K)') +
  ggtitle('Distribution of Hydraulic Conductivity in Different Rock Types')

histogram_plot
```


