---
title: "Using exPrior with real-world data"
author: "Falk Heße"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use WWHYPDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The \pkg{exPrior} packages provides real-world geostatistical data from the World-Wide HYdrological Parameters DAtabank wwwhypda as well as methods for accessing these data using R with the functions \texttt{viewInfo()} and \texttt{getData()}. In this vignette, we are going to exemplify the use of exPrior using these data.

First, we should load the libraries to get access to this function:

```{r}
library(exPrior)
library(ggplot2)
library(gtable)
library(grid)
library(gstat)
```

Let us start by importing data on porosity in sandstone aquifers from the data base.
 
```{r}
exdata <- getData(rockType ='Sandstone', param ='porosity')
```
  
These real-world, ex-situ data can now be used to compute the ex-situ prior using the `exPrior` function (see the vignette simple_example).

```{r}
resExPrior = genExPrior(exdata = exdata, theta = seq(from=0, to=1, by=0.01))
```

Here, the range of the `theta` vector reflects the common-sense intuition that porosity values can only be found between `0` and `1`.

Aftern the completion of `exPrior`, we can visualize both the uninformative and informative distribution of $\theta$ using `plotExPrior`. This function again takes as input the output from `genExPrior` as well as a Boolean asking whether to additionally plot the used data.

```{r}
plotExPrior(resExPrior, plotExData = TRUE)
```

As can be seen, the ex-situ prior distribution for porosity values in sandstone is strongly peaked between `0.2` and `0.3` as determined by the available data. Using this prior can therefore provide a sound foundation for geostatistical inference. 


<!-- # From the vignette generated by default -->

<!-- Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format: -->

<!-- - Never uses retina figures -->
<!-- - Has a smaller default figure size -->
<!-- - Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style -->

<!-- ## Vignette Info -->

<!-- Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette. -->

<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side. -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->

<!-- You can enable figure captions by `fig_caption: yes` in YAML: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         fig_caption: yes -->

<!-- Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**. -->

<!-- ## More Examples -->

<!-- You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`. -->

<!-- ```{r, echo=FALSE, results='asis'} -->
<!-- knitr::kable(head(mtcars, 10)) -->
<!-- ``` -->

<!-- Also a quote using `>`: -->

<!-- > "He who gives up [code] safety for [code] speed deserves neither." -->
<!-- ([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->
