---
title: "Accounting for spatial autocorrelation in ex-situ data"
author: "Falk Heße"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

In most cases, ex-situ data used in assimilation are spatially correlated, since measurements are usually collected in a clustered way. The data assimilation model of exPrior can in principle account for patterns of spatial variability by using multivariate distributions as site-specific distributions. Let us therefore use the following version of the hierarchical model

$$
y_{i,j} \sim \mathcal{N}(\mu_i,\Sigma),
$$
$$
\mu_i \sim \mathcal{N}(\alpha,\tau),
$$
$$
\Sigma = \sigma^2 \exp\bigg( -\frac{h}{\lambda}\bigg),
$$
$$
         (\sigma^2, \lambda, \alpha, \tau) \sim p(\sigma^2)p(\lambda)p(\alpha)p(\tau).
$$

First, we should load the libraries to get access to this function:

```{r}
library(exPrior)
library(ggplot2)
library(gtable)
library(grid)
library(gstat)
```

Unsing this model, we will use spatially correlated, synthtetic data from 3 different sites. Other that that, we will follow the set up of of the vignette "Using exPrior by virtue of a simple example". To assure the sparial correlation in the data, we will first generate them with a geostatistical algorithm from the `gstat` package

```{r}
set.seed(1)
xy <- data.frame("x" = sample(seq(0, 1, 0.01), 22),
                 "y" = sample(seq(0, 1, 0.01), 22))

model = vgm(psill=1, range=1, model='Exp')
g.dummy <- gstat(formula=z~1, locations=~x+y, dummy=TRUE, beta=1, model=model, nmax=20)
exdata_spatial <- predict(g.dummy, newdata=xy, nsim=1)
```

Now, let us create three different subsets of the data in `data`, corresponding to the three ex-situ sites, having somewhat different mean values

```{r}
exdata_spatial[ 1:10, 'sim1'] <- exdata_spatial[ 1:10, 'sim1'] - 3
exdata_spatial[11:15, 'sim1'] <- exdata_spatial[11:15, 'sim1'] - 2.5
exdata_spatial[16:22, 'sim1'] <- exdata_spatial[16:22, 'sim1'] - 3.5
```

correct the column value

```{r}
colnames(exdata_spatial)[3] <- "val"
```

and add the site id.

```{r}
exdata_spatial$site_id = c(rep("S1", 10), rep("S2", 5), rep("S3", 7))
```

The resulting dataframe then looks like this

```{r}
exdata_spatial
```

Now that we have generated our dataframe `exdata_spatial`, containing the ex-situ data together with the spatial coordinates and site_id, we can start inferring the ex-situ prior. Let us start by running `genExPrior` with these data but without accounting for the spatial correlation and look at the uninformative and informative distribution of $\theta$ using `plotExPrior`.

```{r}
theta <- seq(from=-10, to=10, by=0.1)
resExPrior = genExPrior(exdata=exdata_spatial, theta=theta)
plotExPrior(resExPrior, plotExData=TRUE)
```

As can be seen, the results are along the lines of the example in the vignette "Using exPrior by virtue of a simple example". Now let us redo the inference but accounting for the spatial correlations in the data and visualize the results. This is done by setting the flag `spatialCoordinates` to `TRUE`.
 
```{r}
resExPrior_spatial = genExPrior(exdata=exdata_spatial, theta=theta, spatialCoordinates=TRUE)
plotExPrior(resExPrior_spatial, plotExData=TRUE)
```
  
If we compare both ex-situ prior distributions, one without accounting for spatial correlation and one with accounting for it, we see overall similar results. The main difference is that the latter shows a somewhat increased uncertainty, i.e., wider variance. The fact that the more realistic model produces more uncertain results may seem counterintiutive at first. However, the aim of Bayesian inference is not to reduce the uncertainty as much as possible but to correctly represent the uncertainty in the used data and the model.

<!-- # From the vignette generated by default -->

<!-- Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format: -->

<!-- - Never uses retina figures -->
<!-- - Has a smaller default figure size -->
<!-- - Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style -->

<!-- ## Vignette Info -->

<!-- Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette. -->

<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side. -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->

<!-- You can enable figure captions by `fig_caption: yes` in YAML: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         fig_caption: yes -->

<!-- Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**. -->

<!-- ## More Examples -->

<!-- You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`. -->

<!-- ```{r, echo=FALSE, results='asis'} -->
<!-- knitr::kable(head(mtcars, 10)) -->
<!-- ``` -->

<!-- Also a quote using `>`: -->

<!-- > "He who gives up [code] safety for [code] speed deserves neither." -->
<!-- ([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->
